---
title: "Captive Cetaceans: Exploratory Analysis & Mortality Investigation"
author: "Johnny Breen"
date: "27/12/2018"
output: rmarkdown::github_document
---

### Introduction

This is my first ever contribution to the [TidyTuesday](https://github.com/rfordatascience/tidytuesday) project hosted on github. In this post, I am going to analyse a dataset submitted on various cetacean species (incase, like me, you don't know what a 'cetacean' is, it collectively refers to species such as dolphins and whales).

As a student actuary, I am personally quite interested in the study of mortality. Within actuarial work itself, mortality investigations are of course limited to human beings (in the context of selling life insurance and annuities, where setting the appropriate premium is of paramount importance). However, given that this data contains valuable information on the death incidence of captive cetaceans, I'd like to make the focus of this analysis on cetacean *mortality*. 

### Background

According to the [WWF](http://wwf.panda.org/knowledge_hub/endangered_species/cetaceans/threats/bycatch/), one of the leading causes of premature mortality in dolphins is incidental capture (otherwise known as by-catch). If any of you have had the chance to watch the excellent documentary Blackfish, you would have witnessed the incredibly mental torture and suffering imposed on cetaceans by captivity. Cetaceans are designed to travel; [according to conservation biologist Rob Williams](http://www.bbc.co.uk/earth/story/20160310-why-killer-whales-should-not-be-kept-in-captivity), "These are animals that coordinate their movements over scales of tens of kilometres. It's difficult to replicate that in any aquarium". It is estimated that many orcas travel over 62 miles every day so being imprisoned in something like a SeaWorld complex is extremely damaging to the cetacean way of living.

### Preliminary Data Inspection

As a first step, let us load the cetacean data from the TidyTuesdays github repository:

```{r read_csv, warning = FALSE, message = FALSE}
library(tidyverse)
library(magrittr)
library(lubridate)
library(broom)

species_raw <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-12-18/allCetaceanData.csv") %>%
  select(-X1) # X1 appears to be a meaningless extra variable
```

As an initial step, let us investigate the distribution of various categorical variables within the dataset. We will start with the number of different species:

```{r}
# most species are of the bottleneck & orca variety
species_raw %>%
  count(species, sort = TRUE)

# a quick bar plot confirms this - we choose species with a count of above 50
species_raw %>%
  add_count(species) %>%
  filter(n >= 50) %>%
  ggplot(aes(x = species)) +
  geom_bar(width = 0.5) +
  theme_classic() +
  labs(x = "Species",
       y = "Count",
       title = "Approximate distribution of cetacean species",
       subtitle = "The vast majority of cetacean species fall into the Bottlenose type")
```

One important driver of mortality in humans is gender type. It would be interesting to clarify whether this is replicated in cetaceans:

```{r}
# the balance of gender is fairly even - this will make our gender-based estimates more reliable
species_raw %>%
  count(sex)
```

It might also be interesting to investigate how the transition into or out of SeaWorld (notorious for its mistreatment of dolphins - see ['Blackfish'](https://en.wikipedia.org/wiki/Blackfish_(film) ) affects the mortality of dolphins:

```{r}
species_raw %>%
  transmute(seaworld_ind = str_detect(transfers, regex("SeaWorld", ignore_case = TRUE))) %>%
  count(seaworld_ind)
```

### Exploratory Data Analysis

## Mortality Analysis

Let's focus our analysis on deceased species now. Our first step should be to filter the data on 'deceased' species and, in addition to this, engineer the lifespan of each cetacean in the dataset. 

We could focus our analysis solely on cetaceans where the date of birth is known to be accurate, but this would then exclude cetaceans who were born in the wild and subsequently captured - my gut feeling is that this could be an interesting splitting variable so I would like to retain this information. Just be aware that this implies the `lifespan` variable is estimated, in some cases:

```{r}
species_deceased <- species_raw %>%
  filter(status == "Died") %>% 
  mutate(lifespan = as.integer(difftime(statusDate, originDate, units = "days")) / 365.25)
```

So, now, for each species we have a known (or in the case of captured cetaceans) lifespan. I will first narrow down the scope of the fields in the data

```{r}
species_deceased_clean <- species_deceased %>%
  transmute(species, 
            sex, 
            accuracy, 
            acquisition, 
            originDate,
            seaworld_ind = ifelse(str_detect(transfers, regex("SeaWorld", ignore_case = TRUE)), "SeaWorld by-catch", "Other by-catch"), 
            lifespan) %>%
  mutate_if(is.character, as.factor)

species_deceased_clean 
```

Let's first review how the lifespan of captive cetaceans vary across different species types:

```{r message=FALSE}
species_deceased_clean %>%
  group_by(species) %>%
  summarise(avg_lifespan = mean(lifespan, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(species = fct_reorder(species, avg_lifespan)) %>%
  top_n(10, avg_lifespan) %>%
  ggplot(aes(x = avg_lifespan, y = species, colour = species)) +
  geom_point(show.legend = FALSE) +
  geom_segment(aes(x = 0,
                   xend = avg_lifespan,
                   y = species,
                   yend = species),
                   linetype = 'dotted', show.legend = FALSE) +
  expand_limits(x = 0) +
  scale_x_continuous(breaks = seq(0, 35, 5)) +
  theme_classic() +
  labs(x = "Average lifespan (years)",
       y = "Species",
       title = "Analysis of captive cetacean lifespans",
       subtitle = "The Amazon River and Boto deceptively appear to have favourable lifespans")

```

It can be very tempting to draw conclusions from a plot like this; however, it should be noted that we previously inspected the distribution of species available in the data and the *vast* majority of cetaceans fall under the 'Bottlenose' category. This means that the high lifespan of the Amazon River / Boto species is slightly suspicious. We can confirm the distribution of various species by quick inspection of the variable in the deceased data:

```{r}
# top 10 species
species_deceased_clean %>%
  count(species) %>%
  arrange(desc(n))

# bottom 10 species
species_deceased_clean %>%
  count(species) %>%
  arrange(n)
```

Indeed, the top 10 and bottom 10 pulls of the data, as shown above, demonstrate that the representation in each category can be pivotal when drawing conclusions from the data. What might be more instructive is to construct confidence intervals of species lifespan which would take into account the number of data points available in each species type:

```{r}
species_deceased_conf <- species_deceased_clean %>%
  filter(!is.na(lifespan)) %>%
  add_count(species) %>%
  filter(n > 1) %>% # a confidence interval does not make sense for one single data point
  select(species, lifespan) %>%
  nest(lifespan) %>%
  mutate(lifespan_data = map(data, ~ t.test(pull(.)))) %>% # note that 'data' refers to the nested lifespan variable - we're deriving confidence intervals (via the t.test function) for each species
  mutate(lifespan_intervals = map(lifespan_data, ~ tidy(.))) %>% # 'tidy' is from the broom package - it will transform the results of the t.test, for each species, into a corresponding line of data
  unnest(lifespan_intervals) %>%
  transmute(species, avg_lifespan = estimate, lifespan_low = ifelse(conf.low < 0, 0, conf.low), lifespan_high = conf.high) # the lower end of a confidence interval can stray below zero since we are implicitly assuming that each of the species' average lifespan is normally distributed - for sensibility, we set any conf.low values below zero to zero 
```

Now that we have derived a reasonable set of confidence intervals, we can plot the average survival times for each species type once again:

```{r message=FALSE, warning=FALSE}
species_deceased_conf %>%
  top_n(10, avg_lifespan) %>%
  inner_join(count(species_deceased_clean, species)) %>%
  mutate(species = fct_reorder(species, avg_lifespan)) %>%
  ggplot(aes(x = avg_lifespan, y = species, colour = species)) +
  geom_point(show.legend = F) +
  geom_errorbarh(aes(xmin = lifespan_low, xmax = lifespan_high), alpha = 0.75, show.legend = F) +
  theme_classic() +
  scale_x_continuous(breaks = seq(0, 50, 5)) +
  labs(x = "Average lifespan (years)",
       y = "Species",
       title = "Top 10 average cetacean lifespans by species type",
       subtitle = "The confidence interval is unsurprisingly most narrow for the Bottlenose type")
  
```

Now we can see more clearly which estimates are likely to be closer to the true values than others (as is reflected by the width of the confidence intervals). Clearly, we can be quite sure that the Bottlenose lifespan estimate is more or less between around 8 and 10 years. Regarding the False Killer Whale however, the true value could lie anywhere between 2-ish and 14-ish years making the average a little bit meaningless.

## Time-based Analysis

One question I am curious about is how capture rates have changed in prevalence over time:

```{r warning=FALSE, message=FALSE}
# aggregate plot
species_raw %>%
  filter(acquisition == "Capture") %>%
  ggplot(aes(x = year(originDate))) + 
  geom_density() +
  theme_light() +
  labs(title = "Cetacean capture frequency over time",
       subtitle = "What is the cause of the spike post 1970?") 

# by species
species_raw %>%
  filter(acquisition == "Capture") %>%
  mutate(species = fct_lump(species, n = 6)) %>%
  ggplot(aes(x = year(originDate), fill = species)) + 
  geom_density(show.legend = FALSE, alpha = 0.5) + 
  facet_wrap(~species) +
  theme_light() +
  labs(x = "Capture Year",
       y = "Density", 
       title = "Capture frequency split by the top 6 most common species",
       subtitle = "What happened to the 'Spinner'?")
```

According to information on Wikipedia amost half of all spinner dolphins were killed in the 30 years after purse seine fishing for tuna began in the 1950. I wasn't aware of this phenomenon but it could be a potential reason as to why the distribution of the spinner capture rate is so markedly different from the others.

A more general phenomenon to inspect would be how capture / born / rescue rates have changed over time:

```{r}
species_raw %>%
  filter(acquisition %in% c("Capture", "Rescue", "Born")) %>%
  ggplot(aes(x = year(originDate), fill = acquisition)) + 
  geom_histogram(alpha = 0.75, bins = 50) + 
  theme_light() +
  scale_x_continuous(breaks = seq(1940, 2020, 5), minor_breaks = seq(1940, 2020, 5)) +
  labs(x = "Captivity Date",
       y = "Count",
       title = "Histogram of acquisition types",
       subtitle = "The rate of capture appears to drop off rapidly after 1990")
```

Clearly, some legislation has come into practice around the 1990s period which has caused a sudden dropoff in the number of dolphins captured. This is interesting to note because it wouldn't necessarily be anticipated prior to visualisation.

### Survival Modelling

In order to fit a series of survival models to the data we will leverage two additional packages `survival` and `survminer`:

```{r}
library(survival)
library(survminer)
```

This time, we won't be restricting the data to dead species only. Of course, we are interested in calculating the probability of death but in this data we have clear instances of *censoring* present:

* Random censoring is present as certain cetaceans have been released prematurely (a form of censoring that would not have been known to the data collectors at inception)
* Right censoring is present for cetaceans which are currently alive, released or those for which the current status is unknown

These forms of censoring must be incorporated into our model. We will first engineer our survival data in a similar way to how we did it for the deceased species. Before we do so, note that the status of the cetacean is recorded as at 7 May 2017:

```{r}
species_clean <- species_raw %>%
  transmute(species = fct_lump(species, n = 4), # the top four categories have at least 4 observations
            sex, 
            accuracy, 
            acquisition = fct_lump(acquisition, n = 2), # gets rid of 'Stillbirth' and 'miscarriage' 
            seaworld_ind = ifelse(str_detect(transfers, regex("SeaWorld", ignore_case = TRUE)), "SeaWorld by-catch", "Other by-catch"), 
            lifespan = ifelse(status %in% c("Alive", "Released", "Unknown"), as.integer(difftime(as.Date("2017-05-07"), originDate, units = "days")) / 365.25, as.integer(difftime(statusDate, originDate, units = "days")) / 365.25),
            censored = ifelse(status %in% c("Alive", "Released", "Unknown"), 0, 1)) %>%
  mutate_if(is.character, as.factor) %>%
  replace_na(list(seaworld_ind = "Other by-catch"))
```

## Kaplan-Meier Model

Let's first fit a non-parametric model to the data in the form of the Kaplan-Meier survival model. We could split on different variables such as `sex` and `acquisition` but since I plan to fit a cox proportional hazards model shortly (and such a model leverages covariates) I will probably keep this to a minimum. However, I think I will split the data according to whether the cetacean was born in captivity or captured

```{r message=FALSE, warning=FALSE}
species_clean_surv <- Surv(time = pull(species_clean, lifespan), event = pull(species_clean, censored))
species_KM <- survfit(species_clean_surv ~ acquisition, type = "kaplan-meier", data = species_clean)
species_KM %>% 
  tidy() %>%
  mutate(strata = case_when(str_detect(strata, "Born") ~ "Born",
                            str_detect(strata, "Capture") ~ "Captured",
                            TRUE ~ "Other")) %>%
  ggplot(aes(x = time, y = estimate)) + 
  geom_step(aes(colour = strata)) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = strata), alpha = 0.3) + 
  theme_light() +
  facet_wrap(~strata) +  
  scale_x_continuous(breaks = seq(0, 60, 10)) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Time until death (years)",
       y = "Probability of survival",
       title = "Kaplan-Meier Survival Fit",
       subtitle = "According to the fit, less than 50% of born, captive cetaceans are expected to reach 15 years old",
       fill = "Acquisition",
       colour = "Acquisition") 
```

Judging by these models, less than 50% of cetaceans born into captivity make it past the age of approximately 15 years old. This is significantly lower than the life expectancy of cetaceans in the wild where life expectancies typically span the range of 30 to 50 years old. 

Interesting to note is that one would have expected the survival probability of cetaceans born into captivity to be overall lower than those captured (for example, [here](https://us.whales.org/2018/08/23/how-long-do-bottlenose-dolphins-survive-in-captivity/)). However, here we must be careful: we must account for the fact that captured cetaceans live a certain portion of their life *prior* to being captured (captivity) so we should expect the time until death to be, in aggregate, slightly lower for this category of cetaceans.

## Cox Proportional Hazards Model

In this instance, we will use the data from before, but this time we will be able to factor in the effect of different covariates (such as sex and captivity origin) on the probability of survival. A 'hazard' can be thought of as the instantaneous probability of transition from one 'state' to another - in this case, the states are 'Alive' and 'Dead'. The Cox proportional hazards model assumes that:

* The 'hazard' of an individual life is dependent on (i) a *non-parametric* 'baseline' hazard dependent only on time; and (ii) a *parametric* regression-based model, parametrised by a set of time independent factors such as sex
* The ratio of an individual's hazard to the 'baseline' hazard remains in a constant proportion over time

With these assumptions in mind, we can proceed to fit a model to the data:

```{r message=FALSE, warning=FALSE}
species_coxph <- coxph(species_clean_surv ~ species + sex + seaworld_ind, 
                   data = species_clean)

```

We can use the survminer `ggforest` function (it's basically a wrapper around a more complicated ggplot which I do not have the time to create!) to quickly inspect the relative differences in mortality due to the factors we fit the model on above:

```{r message=FALSE, warning=FALSE}
ggforest(species_coxph, data = species_clean, main = "Cetacean Hazard Ratios", refLabel = "Reference")
```

Some notable observations:

* The relative risk of mortality appears to be ~ 27% lower for those cetaceans who have spent time in SeaWorld - this surprises me slightly given what we know about SeaWorld as reported by the media. Nonetheless, it is key to remember that all of these species - regardless of their transit history - are ultimately *captive*. Therefore, whilst there appears to be a relative increase in the probability of survival for those mammals affiliated with SeaWorld, I am fairly certain that by-catch in general is detrimental to the health and condition of any cetacean species. We're simply comparing 'bad' with 'even worse'.
* There is a extremely significant increase in the risk of death amongst species with an unknown gender. This is curious because, as can be observed, there is a non-trivial number of mammals (`N = 105`) contained within this category. Unfortunately, there is no further information on why the gender of these species is unknown.
* The variance of mortality appears to be minimised amongst different species types; however, this could easily be because we omitted any species with a frequency of less than 50 from our analysis. A keen observer will note that the hazard ratios presented above appear to replicate the information communicated in the plot of confidence intervals we made earlier. In truth, we do not have sufficient data on each species to comment conclusively on which species suffer more than others.


