---
title: "Teacher-Student Ratio Analysis"
author: "Johnny Breen"
date: "18/05/2019"
output: github_document
---

# Introduction

In this segment we perform a brief analysis of data on student-teacher ratios across countries worldwide - the data is from the [tidytuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2019/2019-05-07/readme.md) segment on github. 

Limited by time constraints and actuarial exams, this is a relatively short post; however, it showcases some of the functional tools you can use in `purrr` to make your code more powerful. 

# Data import

Let's first read in the data from the github repo:

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
theme_set(theme_light())
# we use 'ts' to refer to 'teacher-student'
ts_ratio_raw <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-07/student_teacher_ratio.csv")

ts_ratio_raw 
```

# Exploratory data analysis

First, let's have a little look at how many countries are contained in the dataset:

```{r}
ts_ratio_raw %>%
  count(country) %>%
  arrange(desc(n))
```

We have 232 'countries' represented - although as you can see some of these are not countries, for example 'World'. 

Next, let's have a look at the representation of different education levels:

```{r}
ts_ratio_raw %>%
  ggplot(aes(x = indicator)) +
  geom_bar(width = 0.5, show.legend = FALSE, aes(fill = indicator)) +
  theme(axis.text.x = element_text(angle = 45, size = 7, hjust = 1)) +
  labs(x = "Education Level",
       y = "Frequency",
       title = "Distribution of education levels")
```

There appears to be a fairly even distribution of different education levels which will be useful if we plan to use this variable as a graphical facetting feature.

Let's have a little look at how the teacher-student ratios correlate (on a linear basis) with time. This will tell us which countries are showing the most consistent changes (good or bad) over time - to do this, we fit a series of linear models to each country in the dataset using `map_*()` functions from the package `purrr`:

```{r}
fit_mod <- function(df, ord = 1) {
  lm(year ~ I(student_ratio^ord), data = df)
}

ts_ratio_rsq <- ts_ratio_raw %>%
  add_count(country) %>%
  filter(n > 10, !is.na(student_ratio), !is.na(year)) %>% # filter on countries with at least 10 data points
  nest(-country) %>%
  mutate(linear_mod = map(data, ~ fit_mod(df = .)),
         linear_mod_summary = map(linear_mod, ~ summary(.)),
         linear_mod_rsq = map_dbl(linear_mod_summary, "r.squared"),
         linear_mod_nobs = map_int(data, ~ nrow(.))) %>%
  arrange(desc(linear_mod_rsq))

# top 15 r-squared values
ts_ratio_rsq %>%
  mutate(country = fct_reorder(country, linear_mod_rsq)) %>%
  top_n(15, linear_mod_rsq) %>%
  ggplot(aes(x = country, y = linear_mod_rsq)) +
  geom_point(aes(colour = country, size = linear_mod_nobs)) +
  expand_limits(y = c(0, 1)) +
  coord_flip() +
  labs(y = "R-squared",
       x = "Country",
       title = "Top 15 most consistent nations",
       subtitle = "Results of regressing student-teacher ratios on time",
       size = "Number of observations") +
  guides(colour = "none")
```

Based on this, Hong Kong appears to be demonstrating a fairly consistent change over time. However, there are some limitations with this approach:

1. We're lumping different education categories together. This is OK if all we care about is aggregate progress but do you remember the `indicator` variable which reflect education levels? Hong Kong, for instance, has two different trajectories for two different education levels:

    ```{r}
  ts_ratio_raw %>%
      filter(str_detect(country, "Hong Kong")) %>%
      ggplot(aes(x = year, y = student_ratio)) +
      geom_point() +
      geom_smooth(method = "lm") +
      facet_wrap(~indicator) +
      labs(x = "Year",
           y = "Student-teacher ratio",
           title = "Student-teacher ratios in Hong Kong spread over time",
           subtitle = "Split by education level shows markedly different progressions")
    ```

2. Our results are based on various volumes of data - we can have a bit more confidence in the Republic of Korea or the Dominican Republic but can't really say much about Aruba at all.

3. We cannot see whether the R-squared value reflects *progresson* or *regression* - i.e. are student-teacher ratios improving in these countries or getting worse?

It is also key to note that the number of observations here does not necessarily reflect the number of years available because some countries have data on a greater number of education levels than others but as at the same year (for instance, primary and secondary education figures in 2012).

Perhaps a better way to represent the data above would be to plot the volatility of student-teacher ratios on the y axis and ratio-year correlations on the x axis. This gives way to a multi-classification of different countries - for instance, a country with a highly negative ratio-year correlation and a high volatility reflects rapid improvement (remember, ratios which *decrease* over time represent increasingly smaller student-teacher ratios so that teachers have less students to cater to):

```{r message=FALSE, warning=FALSE}
ts_ratio_corrs <- ts_ratio_raw %>%
  nest(-c("country", "indicator")) %>%
  mutate(linear_corr = map_dbl(data, ~ cor(.$student_ratio, .$year, use = "pairwise.complete.obs")),
         volatility = map_dbl(data, ~ sd(.$student_ratio, na.rm = TRUE)),
         nobs = map_int(data, ~ nrow(.)))

ts_ratio_corrs %>%
  filter(nobs > 5, !indicator %in% c("World")) %>% 
  ggplot(aes(x = linear_corr, y = volatility)) +
  geom_point(aes(size = nobs, colour = indicator), alpha = 0.75) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_label_repel(data = . %>% filter(abs(linear_corr) > 0.5, volatility > 4.0),
             aes(label = country), size = 2.25) +
  facet_wrap(~indicator) +
  expand_limits(x = c(-1, 1)) +
  guides(colour = "none") +
  labs(x = "Correlation between student-teacher ratio and time",
       y = "Volatility of student-teacher ratios",
       size = "Number of observations",
       title = "Summary of student-teacher ratio volatilities over time",
       subtitle = "Negative correlations and high volatilities are desirable e.g. Gambia") +
  theme(strip.text.x = element_text(size = 6.5))
  
```

Each point in the series of plots above represents a different country. I've higlighted countries which have a suitably high combination of high correlation and volaility. Notice how, for instance, Gambia has a relatively high volatility score as well as a highly negative correlation, suggesting that it is making more rapid improvements than other countries in the area of lower secondary education. This is unsurprisingly backed up by the underlying data:

```{r}
ts_ratio_raw %>%
  filter(country == "Gambia", indicator == "Lower Secondary Education") %>%
  select(education_category = indicator, country, year, student_ratio)
```

The graphical plot also sheds light on poorly performing nations, such as Bhutan in the area of primary education where we can observe a visible increase in student-teacher ratios over time:

```{r}
ts_ratio_raw %>%
  filter(country == "Bhutan", indicator == "Primary Education") %>%
  select(education_category = indicator, country, year, student_ratio)
```

# Conclusion

In this extremely short segment we were able to:

* Identify a set of countries whose student-teacher ratio trajectories were broadly linear, thus identifying the consistently improving nations; and
* Identify which countries are improving fastest and which countries are travelling in the opposite direction

The interested reader could easily augment this analysis by acquiring sovereign data and analysing it in conjunction with the student-teacher ratios presented above - for instance, are these ratios influenced by government spending? Or is there another factor at play here?
