---
title: "Incarceration Rates in the US"
author: "Johnny Breen"
date: "25/03/2019"
output: rmarkdown::github_documents
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 

The data for this post comes from the [Vera institute](https://github.com/vera-institute/incarceration_trends). It comes as a timely reminder of the rampant racial injustice that still exists in the United States prison system. If anyone is looking to learn more about this topic, then I recommend watching the documentary ['13th'](https://en.wikipedia.org/wiki/13th_(film)) - the name refers to the Thirteenth Amendment to the United States Constitution, which abolished slavery throughout the United States and ended involuntary servitude *except* as a punishment for a crime. I also can't help sharing one of my favourite songs from that documentary featuring the excellent Bilal:

<iframe width="560" height="315" src="https://www.youtube.com/watch?v=KO7tVuPHOxA" frameborder="0" allowfullscreen></iframe>

As I am currently facing a relatively busy period at work (and revising for exams), this post will be shorter than usual. I plan to focus on:

* How to build a map using `ggplot2`, `maps` and `mapdata`; and
* How to build an animated plot using `gganimate` (this was my first time too!)

# Read in the data

As always, we first load the packages that we plan to use:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest) # we scrape some information from the web on US state abbreviations
library(magrittr)
library(maps) # contains outlines of various counties, cities and states in the USA
library(mapdata) # same as the above, but a little more detail
library(gganimate) # for animated plots!

theme_set(theme_light()) # sets the theme for all plots (unless we instruct otherwise)
```

Let's first have a quick look at the summary-level data:

```{r message=FALSE, warning=FALSE}
prison_summary_raw <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-22/prison_summary.csv")
```

# Simple plots are effective

The great thing about `ggplot` is that you can plot something non-trivial within a few lines of code. Let's see how the number of individuals in prison (per 100,000) has changed over the course of the last few decades:

```{r}
prison_summary_raw %>%
  ggplot(aes(x = year, y = rate_per_100000, colour = urbanicity)) +
  geom_line() +
  facet_wrap(~pop_category) +
  labs(x = "Year",
       y = "Prison population per 100,000",
       title = "Change in prison populations over time",
       subtitle = "Split by race and gender",
       colour = "Urbanicity") +
  theme_linedraw()
```

Aside from the clear racial discrepancies that are demonstrated by this plot, it is worth noting that the 'Other' category appears to vanish just when the 'Native American', 'Latino' and 'Asian' categories begin to appear - this suggests that the latter categories were likely once grouped as 'Other'. These are the types of useful insights you can gain *instantly* from simply plotting your data: visualisation is a great tool for pattern detection!

# Geospatial analysis

This next section entails my first time to plot geospatial data. Luckily (for me), R comes equipped with several packages to help you to do this, most notably the `maps` package which contains enough data on various large countries (such as the USA and Canada) that you can integrate it with `ggplot2` fairly easily. 

My goal in this section is to develop on the theme we've just explored in the previous plot: how do incarceration rates for African Americans vary across the USA in comparison to other races? First, I'm going to read in the full 'pre-trial' data kindly provided by the Vera institute. This data comprises the incarcerations made on a yearly basis, at county level, from 1970 to 2016:

```{r message=FALSE, warning=FALSE}
pretrial_raw <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-22/pretrial_population.csv")

pretrial_raw
```

One thing you'll notice is that the states are expressed as abbreviations. Unfortunately the `maps` package supplies states as full names. Therefore, I went online to find a table which provides a list of US state abbreviations from [USPS](https://www.usps.com). I've used the `rvest` package written by Hadley Wickham which allows for the extraction of relevant sections from html:

```{r}
us_state_abbrevs <- read_html('https://pe.usps.com/text/pub28/28apb.htm') %>%
  html_node("table.Basic_no_title") %>%
  html_table(header = TRUE) %>%
  as_tibble() %>%
  rename(full_state = `State/Possession`, abbreviation = Abbreviation) %>% # I like lowercase, what can I say?
  mutate(full_state = tolower(full_state))
```

I learnt how to map spatial data in R using the following fantastically (humorous) [link](https://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html). Do take a look if you are curious!

First, let's derive the longitude-latitude data from the `map_data` function in `ggplot2` - we'll then perform an inner join on this table with the abbreviations we derived above. This will join both tables according to their common column name (in this case the full state names) and augment the state information with the relevant abbreviations:

```{r}
us_states <- map_data('state') %>%
  as_tibble() %>%
  inner_join(us_state_abbrevs, by = c("region" = "full_state"))

pretrial_raw %>%
  filter(vars(ends_with("population")) != 0, !is.na(vars(ends_with("population")))) %>%
  mutate(inc_rate = prison_population / population)
```

```{r}
black_region_rates <- prison_mutated %>%
  filter(pop_category == "Black", !is.na(imp_rate)) %>%
  select(-pop_category) %>%
  group_by(State) %>%
  summarise(avg_imp_rate = mean(imp_rate)) %>%
  ungroup()

us_region_rates <- us_states %>%
  left_join(black_region_rates, by = c("region" = "State"))

black_region_high <- us_region_rates %>%
  filter(avg_imp_rate > 0.03) %>%
  distinct(region, .keep_all = TRUE)

# let's first look at one year in time...
ggplot(mapping = aes(fill = avg_imp_rate)) +
  geom_polygon(data = us_region_rates, aes(x = long, y = lat, group = group)) +
  theme_void() +
  geom_label(data = black_region_high,
             aes(x = long, y = lat, label = region)) +
  scale_fill_gradient2(low = "slateblue2", high = "tomato2", midpoint = 0.025, labels = scales::percent_format()) +
  labs(title = "Proportion of african americans in prison from 1970 - 2016",
       subtitle = "The labelled states have the highest black imprisoned population numbers in the US",
       fill = "Avg Jail Proportion")
```

Let's have a look at which states have experienced the most variation over time:

```{r}
black_region_sd <- prison_mutated %>%
  filter(pop_category == "Black", !is.na(imp_rate)) %>%
  select(-pop_category) %>%
  add_count(state, county_name, urbanicity, division) %>% # this will return the number of years available for each observation
  filter(n > 25) %>% # let's stipulate a requirement of more than 25 years of data
  group_by(State) %>%
  summarise(sd_imp_rate = sd(imp_rate), count = n()) %>%
  ungroup()

black_region_sd %>%
  arrange(desc(sd_imp_rate)) %>%
  distinct(State, .keep_all = TRUE) %>%
  select(State, sd_imp_rate) %>%
  top_n(5, sd_imp_rate)
```

Let us see whether that is reflected in an animated plot:

```{r}
by_state_rates <- prison_mutated %>%
  filter(pop_category == "Black", State %in% c("missouri", "tennessee", "texas", "virginia", "michigan")) %>%
  group_by(year, State) %>%
  summarise(avg_imp_rate = mean(imp_rate)) %>%
  ungroup()
  
ggplot(data = by_state_rates, aes(x = year, y = avg_imp_rate, colour = factor(State), group = State)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Year",
       y = "Avg proportion of african americans in jail",
       colour = "State",
       title = "How have imprisonment patterns changed over time?",
       subtitle = "States analysed demonstrated the most variation over the period: 1970 - 2016") +
  transition_reveal(year)
```

Now let's look at all population categories and perform a facet:

```{r}
all_region_rates <- prison_mutated %>%
  filter(!is.na(imp_rate)) %>%
  group_by(State, pop_category) %>%
  summarise(avg_imp_rate = mean(imp_rate)) %>%
  ungroup()

us_region_rates_all <- us_states %>%
  left_join(all_region_rates, by = c("region" = "State"))

us_region_rates_all %>%
  filter(pop_category %in% c("Black", "White", "Latino", "Native American")) %>%
  ggplot(mapping = aes(x = long, y = lat, group = group, fill = avg_imp_rate)) +
  geom_polygon() +
  theme_void() +
  facet_wrap(~pop_category) +
  scale_fill_gradient2(low = "slateblue2", high = "tomato2", midpoint = 0.10, labels = scales::percent_format())

```

### Analysis of prison population split by demographic

```{r}
prison_intvls <- prison_raw %>%
  filter(population != 0) %>%
  transmute(pop_category, inc_rate = prison_population / population) %>%
  filter(!is.na(inc_rate)) %>%
  nest(inc_rate) %>%
  mutate(inc_data = map(data, ~ t.test(pull(.)))) %>%
  mutate(inc_intervals = map(inc_data, ~ tidy(.))) %>%
  unnest(inc_intervals) 

prison_intvls %>%
  mutate(pop_category = fct_reorder(pop_category, estimate)) %>%
  ggplot(aes(y = pop_category, x = estimate)) +
  geom_point(aes(colour = (estimate <= 0.015)), show.legend = FALSE) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, alpha = estimate), show.legend = FALSE) +
  scale_x_continuous(labels = scales::percent_format(),
                     breaks = seq(0, 0.03, 0.005)) +
  labs(x = "Percentage residing in jail",
       y = "Demographic",
       title = "Distribution of jail proportions",
       subtitle = "There is a clear discrepancy in rates between whites and blacks")
```

